\documentclass[12pt]{report}

\usepackage{polyglossia}
\setdefaultlanguage{english}
\setotherlanguage[variant=ancient]{greek}

%\usepackage[style=philosophy-classic, language=english, uniquename=init, mincitenames=2, minbibnames=2]{biblatex}
%\addbibresource{/home/user/.local/share/latex/general.bib}

\title{Climate change and energy transition}
\date{Autumn 2023}
\author{Wester, T.I.}


\renewcommand{\chaptername}{Lecture}


\begin{document}
\maketitle
\tableofcontents

\chapter{}

A central component to climate change is carbon dioxide, which is highly related
to our energy use. Transitioning to a climate friendly global lifestyle, we must
keep in mind the wish to keep our energy budget stable. Energy availability is
tightly correlated with both economic growth and Human Development Index,
neither of which are generally desired to reduce.

When discussing climate problems, there is an intrinsic need for an
interdisciplinary approach. There are for instance questions raised by science
concerning whether we have entered a new geological epoch, how we might get to a
stable and sustainable system regarding climate, and how we might bring justice
to those who greatly damage the climate. However, each of these questions have
normative-, psychological-,  and sociological- dimensions.

\section{The Anthropocene}

The concept of the Anthropocene, the age of man, wherein humans have the
greatest control over the earth, is highly present in the climate change debate.
While the question: ``Have we entered the Anthropocene?'' is a matter of natural
science, the attitude towards it is a philosophical matter.

There are those who fully embrace the Anthropocene, and wish therefore to take
responsibility for the maintenance of the earth.

This approach stands in great contrast one wherein the Anthropocene is despised,
but seen nevertheless as inevitable. This approach typically does not lead to
taking responsibility.

Belatedly, but more responsible, are those who wish to limit the Anthropocene,
mitigating and reducing the negative impact humans have.

\bigskip

There are also other terms proposed against the term ``Anthropocene'', for
instance Homogenocene or Eromocene. Respectively, the age of homogenisation
and the age of loneliness (both referring to biodiversity loss).

\section{Sustainability}

Sustainable development is defined by the WCED as development that meets the
needs of the present without comprising the ability of future generations to
meet their own needs. This definition limits itself greatly to human ability,
and does not properly define the needs current-, and future- generations have.

The assumption is that this ``sustainability'' is indeed possible, partly
because the ethical and political implications of not meeting the needs to the
present are too-repugnant to consider. This belief is not based in empirical
evidence, but rather a matter of necessary hope.

\part{Climate Change}
\chapter{Values and Climate Models}

There is a certain level of uncertainty regarding climate change. There are
various details for which conflicting evidence exists. We do know however that
we are not altogether wrong about anthropocentric climate change. Climate
sceptics almost never pass even the lowest bar for counting as scientific, but
then again, so do many historically-fringe opinions which we now count as fact.

While we know that climate change is largely-, if not primarily- caused by the
emission of greenhouse gasses. The extent to which each affects the global rise
in temperature, is not entirely known. Much like, local effects as opposed to
global ones, various chemical reactions, threshold-theories, and various other
facts.

One may claim that we are to simply adhere to those facts that are exposed to us
through use of the scientific method. However, there are various scientific
methods, whose exact parameters are not rigid. Among scientific models there are
inductive-, deductive-, and abductive- ones, each with their own threshold for
``truth''.

\section{Models}

A large part of the scientific study of climate change is climate modeling. The
aim of a model is to simplify a given environment (called ``the target system'')
to predict the impact of various changes to said environment. Climate models in
particular often combine various models, such as atmospheric-, tectonic-, and
glacial- models, and various complex interactions between them.

We use our models to predict outcomes and inform action, though, due to the fact
that models are -- by definition -- simplified, the modeled outcome rarely
\emph{entirely} conforms to reality.

An ``ensemble model'' is a model which combines the average output of various
models which are calibrated for slightly different purposes. With an ensemble
model, we can project how uncertain we are about a given outcome by analysing
the divergence of each model to the ensemble model.  Using an ensemble model
however relies on certain assumptions, such as all models being equally good in
measuring, and that each of them are scattered around the ``true'' value.

While there are good reasons not to use ensemble models (such a rejecting one of
the assumptions made for using such models), they do show an important matter of
philosophy of science, that being that no hypothesis is every entirely certain.
Meaning that scientists, in accepting theories, make certain value judgements
which lie outside the realm of hard-science.

There is for instance the matter of tolerating false negatives and false
positives. In the case of the climate discussion, false negatives are often
extremely catastrophic, whereas false positives are a matter of wasted funds.
This might lead one to tolerate false positives but not false negatives (or
rather a probability of a given negative being false), but this is a value
judgement which ought to be made by a greater community than merely the
scientific one.

Such a case is argued by Jeffrey, who states that scientists should only express
their expert judgement, not their descriptive claims. Who then is left to make
descriptive claims? This question is not answered to Jeffrey, though a
compelling case can be made to not merely leave this up to the public, nor to a
group working entirely separately from the scientists publishing the
probabilistic facts.

Winsberg argues against Jeffrey and states that scientists can make value
judgements \emph{if} they are transparent about their values. To what extent we
\emph{can} be transparent though is not entirely clear. The value-leadenness of
a model lies in the details, not all of which are obvious to the scientists
themselves, let alone the public,

\chapter{Climate Economics}
\section{Social Discount Rates}
In the case of averting and reverting climate change, the question of cost
dispersal is a big topic. The costs of reversion are high, and the potential
agents capable-, and obliged-, to pay are spatially and temporally dispersed, as
are the victims. Temporal dispersion is particularly problematic, as there is no
straightforward way to assign \emph{current value} to future generations.
Economists have formalized this problem as $\delta + \eta \cdot \gamma = \rho$
where
\begin{tabular}{rcl}
\delta & = & Pure time discount rate\\
\eta & = & Marginal increase of utility per unit of wealth\\
\gamma & = & Elasticity of wealth growth\\
\rho & = & Social discount rate
\end{tabular}

\subsection{Pure time discount rates}
Using this way of calculating the social discount rate, we encounter problems,
largely because this method was intended for economics. For instance, with a
social discount rate of 5\%, one death today would equal a billion deaths 400
years from now, and 5\% is a rather modest discount rate (in the intended
economic context). However, giving no social discount rate, the threat of
eventual extinction -- which, no matter how far off, seems currently inevitable
-- would overrule all current considerations.

In the discussion of social discount rates, there is a distinction to be made
between pure time discounting, and growth rate discounting, with the labels
speaking for themselves.

In the case of time discount rates, when applied to moral considerations, the
question is asked in what way one's position in time is relevant to moral
status. These arguments are particularly popular with utilitarians. In objection
is said that having no time discount rate is unduly demanding for current
generations, as it requires too many sacrifices on our account. To solve this,
utilitarians introduced growth discounting.

\subsection{Growth discount rate}
Since future generations will have more wealth, we ought not pay for them, for
it unjust for the poor to pay the expenses of the rich. This makes it such that
current generations do not need to make great sacrifices, whist still
maintaining that one's position in time is irrelevant to moral consideration.

While this principle on this own allows us to delay action, there are further
considerations which make such delays problematic. It is for instance uncertain
weather certain problem can still be solved in the future, weather future
generations will be wealthier (in every sens of the world, monetarily,
demographically, ecologically, etc.)

Rather than delaying action, one may instead argue for deferring costs. This
would imply that we take action now, but push the cost of action to future
generations. The amount of cost that may be deferred is dependent on the
discount rate, and the damage caused by inaction, and non-deferral. Furthermore,
not deferring costs is politically unrealistic.

\chapter{Sustainability}

Sustainability is not necessarily a matter related to climate change, it is
tightly interwoven with all matters of environmental justice. On its own,
``sustainability'' refers to some ability to perpetuate a certain practice
without being self-destructive. Depending on the area of application, the
particular matter in which sustainability is addressed differs. In commercial
fields, there is often need for a comparison of different sustainable areas,
such as sustainable manufacturing versus sustainable finance. In philosophy,
such comparisons are less important. In both cases, sustainability combines
facts about what is possible with wishes with what is desirable. A further
consideration is the locality of sustainability, where certain practices are
sustainable for individuals, they are not sustainable globally, and vice versa.

All of the above considerations lead to three primary question in the philosophy
of sustainability. Those Being: What should be sustained, what sustainability
dimension we should emphasise, and why is sustainability desirable.

\section{What should we sustain?}

The debate about what ought to be sustained is often discussed in terms of
difference in capital (originating in economics). For instance, we might wish to
sustain natural capital to the detriment of artificial capital. In this debate,
desirability and possibility are very important.

In the debate, two types of sustainability -- the weak-, and strong-conception
-- come forward. Weak sustainability refers to the practice of maintaining total
capital. Replacing one type of capital for another is totally fine as long as
the same functions are performed. The strong conception requires the individual
capital stocks to be maintained. As such, even if all natal functions can be
replaced artificially, a reduction in natural capital is not justifiable. Strong
conceptions of sustainability might allow for exchanges within each capital
stock, such as exchanging animal capital for plant capital because both fall
under natural capital. Stronger versions would not allow this.

\section{Which dimension of sustainability should be emphasised?}

Typically, sustainability is viewed in three pillars, those being: societal-,
environmental-, and economic- sustainability. The question then is what the
proper distribution is for focusing our efforts. Such a three pillar conception
is less popular amongst philosophers of sustainability, though it is prevalent
amongst economists.

Philosophers focus on supply-side versus waste-side sustainability. These are
not necessarily related, and one will not necessarily lead to the other.

Furthermore, depending on which concerns are included in our notion of
sustainability, certain matters become possible of impossible. For instance,
when interspecies considerations are included, certain interpretation make, for
instance the concept of: ``sustainable fishing'' conceptually impossible.

\section{For which purpose?}

Sustainability ultimately comes down to human well-being. Well being must be
sustained so that future generations have the ability to meet their needs,
preferably just as well as we do, or better. Here too, there is a difference
between economists and philosophers, as sustainable economic growth is seen my
many philosophers as an oxymoron. Philosopher focus instead on human
development, though this matter is also tightly linked with economic growth.

Typically, the sustainability debate becomes one of theories of justice, though
with an additional ecological angle. The theories of justice considered
typically include intergenerational, global, and interspecies concerns.
Alongside the aforementioned ecological concerns, these theories are often
called ``planetary justice''.

\chapter{Democracy}

A decision-making-process can be called ``democratic'' if it derives its
legitimacy from the people who are affected by the decision. As a collective
depiction-making-process, democracy essentially entails a kind of equality of
decision makers at the vital stage of the process.

Democracy is not desirable under every circumstance, and not every variant of
Democracy is equally desirable either. There is for instance often a trade-off
between directness and effectiveness, where direct democracy is seen as too time
consuming to be effective in large states. This leads into the discussion
concerning who should be included in the decision-making-process.

In the case of environmental concerns,there are two major approaches to this
latter question. The first is the cosmopolitan answer, which bases its approach
on the all-affected-interests answer, which leads to a globalist approach.
Alternatively, there is the eco-anarchist response, which holds that the
political community has to be local, and that -- in the case of environmental
concerns -- each ecosystem requires its own tailored approach and therefore a
local governmental body.

\section{Desirability}

According to Jean-Jacques Rousseau, there is a difference between individual
interests and general will. People ought to base their vote on what is best for
society. Through democracy therefore, we get insight into the general will,
therefore leading us to desirable outcomes. Furthermore, through the democratic
process, we gain moral-, and civil-freedom through self-governance.

In modern times, the so-called ``instrumental justification'' is used. This does
away with Rousseau's idea that democracy produces good depictions, instead
claiming that it produces better depictions than any other process. Furthermore,
through engaging in the democratic process, we develop a stronger moral
character.

Both of these accounts are littered with holes, as there is no obvious way in
which the epistemic state of the individuals of a state are related to justice
and good outcomes. This problem is reinforced by populism as well as the
complexity of issues and of the democratic processes itself. 

Democracy is also argued for due to its intrinsic values, such as quality,
fairness, and freedom. This does not change the fact that minority opinions are
rarely represented in democratic systems, and there are also various simply
strong opinions represented to various extents. Resolving these problems goes
against the intrinsic value of equality.


\section{Ecological democracy approaches}

A frequently proposed solution in environmental literature to the problems with
democracy is the ideal of deliberative democracy. In a deliberative democracy
the engagement with- and discussion of- ideas is of utmost importance.
Nevertheless, democracy -- of any form -- cannot guarantee the realisation of
adequate climate policies. As such, there are now growing voices for aristocracy
where an eco-authoritarian body decides on climate policies. Such solutions
however face the issues of cooperation and legitimacy.

\chapter{Compound problems of climate justice}

Climate change is, by all accounts, a moral problem, but exactly how and where
the moral dimensions of climate issues lie remains a question with many
obscuring factors. Some of these confounding actors have already been covered in
this course, such as intergenerational-, international-, and (to a lesser extent
discussed) interspecies- considerations.

One major further problem comes from the \emph{global} element in ``global
warming''. This element entails that global action is necessary. One primary
concern for the global community is where to focus our collective efforts, the
primary options being: mitigation, adaptation and damage control. In the worst
case scenario, this discussion leads to a lack of action in every area.

Collective action problems in general result from diversity in action for
individuals, or -- in this case -- individual states. This diversity in action
arises through various mechanisms, for instance through a lack of knowledge.
This lack could be inherent, for instance when research on the matter at hand is
inconclusive, ongoing, or simply not done. Though the lack of knowledge can also
come from simple ignorance, willful or otherwise. There is also the matter of
individual restrictions, where action is much more costly for some individuals
than others. Lastly, there is also the matter of rational self-interest where,
in a prisoners-dilemma like-scenario, it is better for a given individual not to
cooperate \emph{given} that others do cooperate.  Climate change is of course a
matter of collective action because it is concerned with a public good.
Fundamentally though, a stable climate is not optional, but rather an essential
public good, thus making action a moral \emph{demand}, rather than a moral
\emph{desire}. Nevertheless, one can enjoy a stable climate without contributing
to its maintenance.

The non-identity problems, raised by Parfit, was (and partially still is) a
large part of the intergenerational concerns for climate change. Currently, this
matter is often addressed in manners which still leave us with a moral
obligation to future generations. Approaches include historical justification,
generational overlap, and earth stewardship approaches. All of these approaches
allow for the central claim of the non-identity problem, namely that we cannot
have obligations for non-existing agents, because they might never come into
existence, and yet require us to take action. The method that departs most from
classical intergenerational approaches is the so-called stewardship approach,
which postulates that we have a duty to the earth itself, rather than the future
generations that live on it. The earth, as an enduring patient, carries our
responsibilities to future generations. The historical justification points out
that we do in fact feel and observe that humans have concern for future
generations, even if this is not expressible in the analytic philosophical
system. The generational overlap approach uses the fact that many generations
live at one, with each having responsibilities for the rest. As such, we have no
responsibility to unborn generations (as is claimed by the non-identity
problem), but we do have obligations for the young and vulnerable. This
considerations allows for the forming of a chain which carries through all the
way to generations in the far future.

\chapter{Cultural-, and interspecies- considerations}

Climate burdens go beyond monetary considerations as we have various mitigation
and adaptation burdens which results in certain losses being exchanged for
certain wins. For instance, as we transition to cleaner energy sources, and
traditional jobs are lost, not only are people's well being affected monetarily,
this also affects culture and identity. Likewise, culture and identity will be
diminished by a lack of climate action.

In the section above, four different angles are already discussed. Concisely
these are:
\begin{enumerate}
	\item
		Mitigation burdens
	\item
		Adaptation burdens
	\item
		Impacts of climate change when mitigation is lacking
	\item
		Implants of climate change when adaptation is lacking.
\end{enumerate}
All climate concerns will likely include several of these dimensions.

Claire Heyward writes about these issues, though her analysis is largely limited
to cultural loss in the case of low-emitting nations. Focusing not on cultural
change or adaptation, and not on losses of high-emitting nations.


\textit{The latter half of this lecture was on interspecies considerations, I
was unable to attend this part. See Plamer's text}

\part{Energy Transition}
\chapter{Justice and Energy}

While climate change is ongoing with ever increasing risks, and greater
projected consequences to a lack of mitigation. A major part of our negative
climate contribution is the emission of atmospheric carbon dioxide due to the
burning of fossil fuels. Since the energy availability that results from the
burning of these fuels leads to an overall higher state of well-being,
transitioning to a new form of energy to mitigate climate change, must not lower
this level of well-being through a reduction of availability.

Energy ethics and energy justice have historically taken second place, given
less attention in comparison to climate justice and climate change. Despite
this, the energy transition ahead of us is by par the largest, most
unprecedented transition thus far, with us both using more energy than ever and
more fossil fuels than ever.

Energy injustice is furthermore already present. Many people lack aces to
electricity and many more depend on solid fuels, thus drastically lowering their
standard of living compared to countries where electricity is pervasive. Energy
production also is highly correlated with human rights abuses such as poor
working conditions and forced labour.

\section{Midway ethics}

Midway ethics is an approach towards moral dilemmas where we do both historical
and ethical analysis, thus energising empirical science and moral theory. This
position is developed because neither self-denying idealism (qua Rawls) is not a
convincing stance, though neither is self-centered realism (empirical answer).

This system does borrow a lot from both of these unsatisfactory positions, such
as Rawlsian reflective equilibrium. This is a system where we reflect on
principled, adjust them based on new information, and reflect again.

\chapter{Carbon capture and Storage}

Carbon capture and storage\footnote{henceforth CCS} is heralded as a potentially
game chaining technology, allowing for the mitigation-, or complete aversion- of
climate change whilst allowing the continuation of business as usual. Despite
its great promises, CCS has many problems facing its implementation including
placement, financing, and cultural impact (or lack thereof).  These roadblocks
often lead to considerations of justice.

\section{Implementation justice}

The implementation of CCS technologies by developed countries allows undeveloped
nations to have access to cheap energy, of instance through coal plants, while
mitigating their climate impact. This addresses the concern of injustice in
disallowing developing nations to develop, as aces to cheap and abundant energy
seems to be a prerequisite for national development.

When talking about resource depletion however, the implementation of CCS would
accelerate the problems regarding this. CCS facilities, both in capture,
transport, and storage require energy. Therefore, for every four current coal
powered plants, we would need to build a fifth just to power the CCS facility.
Thus causing a 25\% increase in resource depletion.

Furthermore, there are limited sites where we could store carbon underground,
requiring a porous rock capped by a dense rock to keep the carbon from leaking
out. These sites are also a scarce resource which will run out.

Furthermore, the location of both capture and storage facilities raises issues
of environmental degradation, both in terms of biodiversity loss and ascetic
considerations.

\section{Technological justice}

There are also risks associated more closely with the technology itself.

There are for instance risks of leakage associated with both transport and
storage of carbon dioxide. To what extent we wish to accept those risks is a
matter of justice. After all, complete inaction is extremely risky, though
complete inaction is not an option whatsoever. Compared then to other possible
mitigation strategies, for instance how we measure the risks of implementing CCS
with chance of leaking in comparison to implementing fully renewable energy
production with associated risks, is not a straightforward matter.

Furthermore, most CCS technologies available now use a lot of water in the
capture process, which is a resource not captured within mere-carbon-thinking.
These technologies are furthermore protected under patents with the aim to
financially incentivise innovation, though in practice it also reduced free and
open collaboration and iterative improvement.

There is also the matter on maintenance. CCS storage facilities in particular
need to be maintained for centuries if we wish to prevent leakage. This commits
future generations to fix our problems in a manner not too different from
requiring them to mitigate climate change directly. This discussion is often
accompanied with the maintenance costs of nuclear-waste storage and associated
risks.

\section{Responsibility}

There is also the question of \emph{who} ought to implement and finance CCS. As
mentioned earlier, there is a case to be made that developed nations ought to
take this burden. Though there are also those arguing that the main polluters
ought to pay, be that current polluters or historical ones.

There are also discussions regarding whether it is even morally good or morally
required to invest in CCS. This is often based on a rights based perspective,
where, due to the scarcity of resources, perhaps we ought not invest in CCS, but
rather in renewable energy, as investment in CCS diverts financial assets away
from morally good causes.

\section{Inaction}

One major problem with CCS and all other negative emission
technologies\footnote{Henceforth NET} is the problem of inaction. In short, NET
allows countries and companies to overshoot their carbon budgets as long as they
have the crutch of CCS and NET. Worries are that large scale implementation of
CCS would stifle innovation in renewable technologies. Furthermore, as CCS
becomes cheaper, as emerging technologies tend to do, they will further increase
the financial attractiveness of coal pants. This still leaves us with the
problems of resource depletion, overconsumption, and non-carbon environmental
degradation.

\textit{Prosperity without Growth; Tim Jackson}

\chapter{Nuclear energy and the far future}

Nuclear energy is a core part of the energy transition discussion. Back before
nuclear energy was widely developed -- and critically -- before the major
accidents such as Chernobyl and Fukushima, nuclear energy was seen as a better,
alternative to renewable such as wind and solar power due to the latter's low
density. Currently, the discussion sill focuses on these aspects, though a core
of the discussion is also the idea that nuclear energy may be necessary -- at
least as a transition technology -- if we wish to advert climate change.

The risks of Nuclear facilities, aside from meltdown accidents, also include
nuclear waste storage. Due to the long lived nature of depleted uranium, storage
facilities need to withstand centuries of corrosion and other changes. This
matter also raises the issue of how to warn sentient beings in the far future
about the dangers of these facilities.

Dangers are of course not exclusive to nuclear reactors. There are even those
who view climate change as one of the dangers of fossil fuel plants. Nuclear
every must therefore be viewed in comparison with other energy technologies, and
the risks of not having enough energy available.

Through the combination of all of these factors, the temporal dimension of
Nuclear power considerations is unprecedentedly long. Due to this lack of
precedent, our ethical-, and economic- theories are not well suited to address
Nuclear Energy. This is influenced furthermore by the fact that the threat of
climate change is both largely nebulous and far into the future.

\section{sustainability}

Depending on our definition of sustainability, Nuclear energy fits the
definition to a greater or lesser extent. Critically, nuclear energy is largely
sustainable when it comes to atmospheric carbon dioxide, though it is not when
it comes to natural resource depletion. These two angles are incommensurable,
meaning that one cannot translate resource depletion into a carbon-equivalent.

Sustainability addresses the question of the needs of future generations.
Determining those needs is not a trivial matter however, especially when
addressing the time frame over which nuclear energy is relevant. As the
``needs'' can be defined by each individual, we do not want to limit the needs
of future generations by limiting things which we do not consider needs, but
they might.

\end{document}
